```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

\newcommand{\ba}{\mathbf{a}}
\newcommand{\bc}{\mathbf{c}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bz}{\mathbf{z}}
\newcommand{\bA}{\mathbf{A}}
\newcommand{\bB}{\mathbf{B}}
\newcommand{\bC}{\mathbf{C}}
\newcommand{\bD}{\mathbf{D}}
\newcommand{\bI}{\mathbf{I}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\bM}{\mathbf{M}}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bW}{\mathbf{W}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bY}{\mathbf{Y}}
\newcommand{\dC}{\mathbb{C}}
\newcommand{\dE}{\mathbb{E}}
\newcommand{\dM}{\mathbb{M}}
\newcommand{\dP}{\mathbb{P}}
\newcommand{\dV}{\mathbb{V}}
\newcommand{\cH}{\mathcal{H}}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cR}{\mathcal{R}}
\newcommand{\dR}{\mathbb{R}}
\newcommand{\bgb}{\boldsymbol{\beta}}
\newcommand{\bga}{\boldsymbol{\alpha}}
\newcommand{\bgt}{\boldsymbol{\theta}}
\newcommand{\bgS}{\boldsymbol{\Sigma}}
\newcommand{\bzero}{\mathbf{0}}
\newcommand{\bones}{\mathbf{1}}
\newcommand{\hy}{\hat{y}}
\newcommand{\he}{\hat{e}}
\newcommand{\hgb}{\hat{\beta}}
\newcommand{\hby}{\hat{\mathbf{y}}}
\newcommand{\hbe}{\hat{\mathbf{e}}}
\newcommand{\sumin}{\sum_{i=1}^n}
\newcommand{\sumjn}{\sum_{j=1}^n}

# Intro to R

These notes introduce core R concepts you’ll use throughout the course.
Each section includes short explanations and runnable code.

> Tip: You can run code line-by-line in RStudio with **Ctrl+Enter** (Windows) / **Cmd+Enter** (Mac).

## Basic R Operations and Concepts

R works like a powerful calculator *and* a programming language.
You type commands into the **Console**, and R evaluates them.

### The Console and Scripts

- **Console**: where commands run immediately.
- **Script (.R)** / **R Markdown (.Rmd)**: where you write reproducible work you can save and re-run.

### Comments

Use `#` to write comments. R ignores them.

```{r}
# This is a comment
2 + 2
```

### Printing and Output

Typing an object name prints it. You can also use `print()`.

```{r}
x <- 10
x
print(x)
```

## Arithmetic

R supports standard arithmetic operators:

- Addition: `+`
- Subtraction: `-`
- Multiplication: `*`
- Division: `/`
- Exponentiation: `^`
- Integer division: `%/%`
- Remainder (mod): `%%`

```{r}
# Basic arithmetic
5 + 3
10 - 4
6 * 7
20 / 5

# Powers
2^5

# Integer division and remainder
17 %/% 3
17 %% 3
```

### Order of Operations

R follows standard order of operations. Use parentheses to be explicit.

```{r}
3 + 2 * 5      # multiplication first
(3 + 2) * 5    # parentheses first
```

### Logical Comparisons

These return `TRUE` or `FALSE`.

```{r}
5 > 3
5 == 3
5 != 3
5 >= 5
```

## Assignment, Object Names, and Data Types

### Assignment

Use `<-` (common in R code) or `=`.

```{r}
a <- 5
b = 12
a + b
```

### Object Names

Good names are readable and informative.

**Valid examples**

- `height`
- `exam_score`
- `x1`

**Avoid**

- spaces (`exam score`)
- starting with numbers (`1st`)
- reserved words (`if`, `for`, `TRUE`, `FALSE`)

```{r}
exam_score <- 92
final_grade <- "A"
```

### Common Data Types

#### Numeric (double)

```{r}
x <- 3.14
typeof(x)
class(x)
```

#### Integer

```{r}
y <- 7L
typeof(y)
class(y)
```

#### Character (strings)

```{r}
name <- "Rene"
typeof(name)
class(name)
```

#### Logical

```{r}
flag <- TRUE
typeof(flag)
class(flag)
```

### 3.4 Missing Values

R uses `NA` for missing data.

```{r}
z <- c(1, NA, 3)
z
is.na(z)
```

`NA` affects calculations unless you remove missing values.

```{r}
mean(z)                 # returns NA
mean(z, na.rm = TRUE)   # removes NA before computing mean
```

## Vectors

Vectors are the **basic building block** in R: an ordered collection of values.

### Creating Vectors with `c()`

```{r}
v <- c(2, 4, 6, 8)
v
```

### Vector Length

```{r}
length(v)
```

### Vector Arithmetic (Vectorized Operations)

Operations apply element-by-element.

```{r}
v + 1
v * 2
v^2
```

### Sequences with `:` and `seq()`

```{r}
1:10
seq(from = 0, to = 1, by = 0.2)
seq(from = 1, to = 10, length.out = 5)
```

### Repetition with `rep()`

```{r}
rep(5, times = 4)
rep(c("A","B"), times = 3)
rep(1:3, each = 2)
```

### Indexing Vectors

#### By position

```{r}
v <- c(10, 20, 30, 40, 50)
v[1]      # first element
v[3]      # third element
v[c(2,5)] # second and fifth
```

#### By negative indexing (remove elements)

```{r}
v[-1]     # all but first
v[-c(2,4)]
```

#### By logical indexing

```{r}
v[v > 25]     # keep values greater than 25
v[v == 20]    # values equal to 20
```

### Named Vectors

Names make code easier to read.

```{r}
grades <- c(Midterm = 88, Final = 93, Project = 90)
grades
grades["Final"]
```

## Functions and Expressions

### Functions

R has many built-in functions: `mean()`, `sum()`, `sd()`, etc.

```{r}
x <- c(2, 4, 6, 8, 10)

sum(x)
mean(x)
sd(x)
min(x)
max(x)
```

#### Function arguments

Arguments control how the function behaves.

```{r}
round(3.14159, digits = 2)
mean(c(1, NA, 3), na.rm = TRUE)
```

### Expressions and Nesting

You can combine (nest) functions.

```{r}
x <- c(1, 2, 3, 4, 5)
sqrt(sum(x^2))
```

### Creating Your Own Function

Use `function()` to define reusable code.

```{r}
# A simple function: compute z-scores
zscore <- function(x) {
  (x - mean(x)) / sd(x)
}

zscore(c(10, 12, 15, 20))
```

#### Another example: a summary function

```{r}
quick_summary <- function(x) {
  c(
    n = length(x),
    mean = mean(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE)
  )
}

quick_summary(c(1, 2, 3, NA, 5))
```

## Getting Help

R has excellent built-in help tools.

### Help Pages

Use `?` or `help()`.

```{r}
?mean
help(sd)
```

### Examples in Help Files

Many help pages include examples you can run:

```{r}
example(mean)
```

### Searching for a Function

If you don’t know the exact name, use `help.search()` or `??`.

```{r}
??regression
help.search("histogram")
```

### Inspecting Objects

```{r}
x <- rnorm(5)
str(x)
class(x)
typeof(x)

ls()        # list objects in your environment
rm(x)       # remove x
ls()
```

### Getting Package Help

If you use a package, load it first (if installed).

```{r}
# install.packages("ggplot2")  # run once if needed
library(ggplot2)

?ggplot
```

