# Data Description

In this section, we explore how data can be **classified**, **summarized**, and **interpreted**.  
We introduce key concepts for both **categorical** and **quantitative** data, and we examine graphical and numerical tools used in statistics.

---

## Data Simulation

For this section we will simulate a data set for student performance in math and verbal tests accross 10 schools. For each student we will include the following variables:

Variables:

- Student ID
- Student School
- Student Grade
- Student Age
- Income Level
- Education Level Parents
- Math Score
- Verbal Score

```{r school-data-simulation, echo=FALSE}
# Data Simulation
set.seed(2026)

# Simulation Parameters
numSch    <- 10
namSch    <- paste0("School ", 1:numSch)
numGra    <- 9
numStuGra <- 100
numStu    <- numSch * numGra * numStuGra
minScoMat <- 0
maxScoMat <- 800
minScoVer <- 0
maxScoVer <- 800
namIncLev <- c("Low", "Middle", "High")
namEduPar <- c("No Highschool", "Highschool", "College", "Graduate School")

# Student School
sch    <- sample(namSch, size = numStu, replace = TRUE)
gra    <- sample(1:numGra, size = numStu, replace = TRUE)
age    <- round(gra + 5 + runif(n =numStu, min = -0.6, max = 0.6))
incLev <- sample(namIncLev, size = numStu, replace = TRUE, prob = c(0.3, 0.6, 0.1))
eduPar <- sample(namEduPar, size = numStu, replace = TRUE)

# Simulate Scores
matSco <- numeric(length = numStu)
verSco <- numeric(length = numStu)
numSel <- c()
for(i in 1:numSch){
  # School Performance Level
  schPer <- runif(n = 1, min = 0, max = 1)
  # Selects School
  selSch <- namSch[i] == sch
  for(j in 1:length(namIncLev)){
    # Income Performance
    incPer <- j / length(namIncLev)
    # Selects Income
    selInc <- namIncLev[j] == incLev
    for(k in 1:length(namEduPar)){
      # Parent Education Performance
      eduPer <- k / length(namEduPar)
      # Selects Parent Education Level
      selPar <- namEduPar[k] == eduPar
      for(l in 1:numGra){
        # Grade Level Performance
        graPer <- l / length(numGra)
        # Selects Grade Level
        selGra <- l == gra
        # Selected Students
        sel         <- selSch & selInc & selPar & selGra
        numSel      <- c(numSel, sum(sel))
        numStuClu   <- sum(sel)
        matSco[sel] <- 0.2 * schPer + 0.1 * incPer + 0.1 * eduPer + 0.6 * graPer
        matSco[sel] <- matSco[sel] + rnorm(n = numStuClu, mean = 0, sd = 0.2)
        verSco[sel] <- 0.1 * schPer + 0.1 * incPer + 0.1 * eduPer + 0.7 * graPer
        verSco[sel] <- matSco[sel] + rnorm(n = numStuClu, mean = 0, sd = 0.1) 
      }
    }
  }
}

# Transforms Scores to 0-800 Scale
matSco <- (matSco - min(matSco)) / (max(matSco) - min(matSco)) * 850
verSco <- (verSco - min(verSco)) / (max(verSco) - min(verSco)) * 850
matSco[matSco > 800] <- 800
verSco[verSco > 800] <- 800
matSco <- round(matSco)
verSco <- round(verSco)

# Creates a data frame
schDat <- as.data.frame(1:numStu)
schDat <- cbind(schDat, sch, gra, age, incLev, eduPar, matSco, verSco)
colnames(schDat) <- c("Student_ID", "School", "Grade", "Age", "Income_Level", "Parent_Education", "Math_Score", "Verbal_Score")
```

## Types of Data

### Quantitative vs. Qualitative Data

```{definition, data-types, name="Types of Data"}
Data can be classified into **quantitative** (numerical) and **qualitative** (categorical) types.  
Quantitative data consist of numerical measurements; qualitative data describe categories or attributes.
```

### Quantitative Data

Quantitative data represent **numerical values** that can be measured.

Examples:

*   Heights of students
*   Weights of packages
*   Exam scores

They can be:

*   **Discrete**: countable (e.g., number of children)
*   **Continuous**: measured on a continuum (e.g., height, time)

### Qualitative Data

Qualitative (categorical) data represent **non‑numeric labels or groups**.

Examples:

*   Sex
*   Favorite color
*   Yes/No responses
*   Satisfaction Level

These categories may or may not have a natural order.

---

## Why Do We Have to Describe Data?

Describing data allows us to:

*   Explain data.
*   Identify **patterns** or **trends**
*   Summarize large datasets compactly
*   Compare groups or populations
*   Detect **outliers** or unusual observations
*   Prepare for inferential analysis

Raw data alone rarely reveal useful insight without proper summarization.

---

## Types of Data Description

### Numerical Description

Focuses on numerical summaries such as:

*   Mean
*   Median
*   Standard deviation
*   Quantiles

### Graphical Description

Focuses on visualizations such as:

*   Histograms
*   Bar charts
*   Pie charts
*   Boxplots

Both approaches are essential for understanding data behavior.

---

## Techniques to Describe Categorical Data

### Graphical Techniques

#### Bar Charts

Bar charts display the frequency of each category using rectangular bars.

Example in R:

```{r barplot-schools, }
barplot(table(schDat$School), col="steelblue")
```

### Pie Charts

Pie charts show proportions as slices of a circle.


```{r pichart-income}
pie(table(schDat$Income_Level))
```

### Numerical Techniques

The main numerical summaries for categorical data include:

*   **Counts** (frequencies)
*   **Proportions**
*   **Relative frequencies**

Example:

```{r table-income}
# Frequency Table
table(schDat$Income_Level)
# Proportional Table
prop.table(table(schDat$Income_Level))
```

---

## Techniques to Describe Quantitative Data

### Displaying All the Data

Tools:

*   Raw data lists
*   Dotplots
*   Stem‑and‑leaf plots

Example:

```{r dat-des-stem-and-leaf-plot}
stem(c(23,25,25,30,31,33,35,40))
```

---

## Frequency Table

A frequency table groups quantitative data into **intervals** (classes).

### Class Frequency

```{definition, class-frequency, name="Class Frequency"}
The **class frequency** is the number of observations that fall within a specific interval.
```

### Relative Frequency

```{definition, rel-frequency, name="Relative Frequency"}
The **relative frequency** is the class frequency divided by the total number of observations.
```

Example:

```{r frequency-table}
# Frequency Classes
breaks    <- seq(0, 800, by = 100)
matScoBin <- cut(matSco, breaks = breaks, right = TRUE) 
table(matScoBin)
prop.table(table(matScoBin))
```

---

### Histograms

Histograms partition quantitative data into **bins** and display frequencies as bars.

#### How to Build Them

Steps:

1.  Choose number of intervals (bins).
2.  Determine bin width.
3.  Count frequencies per bin.
4.  Draw adjacent bars.

#### Number of Intervals

Common guidelines:

*   Square-root rule: $$k = \sqrt{n}$$
*   Sturges' rule: $$k = 1 + \log_2(n)$$

Example:

```{r histogram-schools}
hist(schDat$Math_Score[schDat$Grade == 1], col="skyblue", main = "Histogram of Math Scores for First Grade", xlab = "Math Scores")
```

#### Characteristics of Histograms

##### Unimodal

One peak.

##### Bimodal

Two distinct peaks.

```{r histogram-schools-bimodal}
hist(schDat$Math_Score[schDat$Grade == 3 | schDat$Grade == 5], col="skyblue", main = "Histogram of Math Scores for Third and Fifth Grade", xlab = "Math Scores", breaks = 20)
```

##### Symmetry

*   **Symmetric**: left and right sides mirror
*   **Skewed Right**: long right tail
*   **Skewed Left**: long left tail

---

## General Guidelines for Successful Graphics

*   Label axes clearly
*   Use consistent scales
*   Avoid unnecessary decoration
*   Ensure readability
*   Highlight key trends

---

## Measures of Central Tendency

### Mode

```{definition, mode-def, name="Mode"}
The **mode** is the value that occurs most frequently in a dataset.
```

### Median

```{definition, median-def, name="Median"}
The **median** is the middle value when data are ordered.
```

### Mean

```{definition, mean-def, name="Mean"}
The **mean** is the arithmetic average of all observations $x_1, x_2, \ldots x_n$. Denoted by:
  $$ \bar{x} = \frac{\sum_{i=1}^n x_i}{n} $$
```

Example:

```{r dat-des-mean-median-comparisson}
mean(schDat$Math_Score[schDat$Grade == 1]); median(schDat$Math_Score[schDat$Grade == 1])
```

---

## Measures of Variability

### Range

```{definition, range-def, name="Range"}
The **range** is the difference between the maximum and minimum values in a dataset.
```

```{r dat-des-min-max-range}
c(max(schDat$Math_Score[schDat$Grade == 3]),
min(schDat$Math_Score[schDat$Grade == 3]))
max(schDat$Math_Score[schDat$Grade == 3]) - min(schDat$Math_Score[schDat$Grade == 3])
# Average score 3rd Grade
mean(schDat$Math_Score[schDat$Grade == 3])
# Average score 4th Grade
mean(schDat$Math_Score[schDat$Grade == 4])
```

### Quantiles

Quantiles divide data into equal-sized groups.

### Quantile Plot

```{r quantile-plot_math_scores}
plot(sort(schDat$Math_Score[schDat$Age == 11]),
     ppoints(length(schDat$Math_Score[schDat$Age == 11])),
     xlab = "Math Score",
     ylab = "Quantile")
```

### Percentiles

The p-th percentile is the value below which p% of the data fall.

### Quartiles

Quartiles divide data into four equal parts:

*   Q1
*   Q2 (median)
*   Q3

### Deviation

A deviation is the difference between an observation and the mean. That is
$$ x_i - \bar{x} $$

### Variance

```{definition, variance-def, name="Variance"}
The **variance** measures the average squared deviation from the mean.
$$ s^2 = \frac{\sum_{i=1}^n (x_i - \bar{x})^2}{n-1} $$
```

### Standard Deviation

```{definition, sd-def, name="Standard Deviation"}
The **standard deviation** is the square root of the variance.
```

Example:

```{r sd-and-variance-math-scores}
sd(schDat$Math_Score)
var(schDat$Math_Score)
```

### Absolute Deviation

Average of absolute deviations from the mean.

### Median Absolute Deviation (MAD)

Robust measure of variability.

```{r median-absolute-deviation-math-scores}
mad(matSco)
```

---

### 68%, 95%, and 99.7% Empirical Rule

For bell‑shaped distributions:

*   68% of values fall within 1 SD of mean
*   95% within 2 SD
*   99.7% within 3 SD

---

### Coefficient of Variation

The CV measures spread relative to the mean:

$$
CV = \frac{SD}{\text{mean}}
$$

---

### The Boxplot

Displays:

*   Median
*   Quartiles
*   Whiskers
*   Outliers

Example:

```{r boxplot-math-scores}
boxplot(schDat$Math_Score[schDat$Grade == 3], schDat$Math_Score[schDat$Grade == 4], schDat$Math_Score[schDat$Grade == 5])
```

---

### Outliers

Observations unusually far from the rest:

*   Often defined as:
    *   Below Q1 – 1.5(IQR)
    *   Above Q3 + 1.5(IQR)

---

## Techniques for Two Variables

### Contingency Tables

Summarize two categorical variables.

```{r table-income-education}
table(schDat$Income_Level, schDat$Parent_Education)
```

### Stacked Bar Graph

```{r dat-des-stacked-graph}
barplot(table(schDat$Income_Level, schDat$Parent_Education), beside=FALSE)
```

### Cluster Bar Graph

```{r dat-des-cluster-graph}
barplot(table(schDat$Income_Level, schDat$Parent_Education), beside=TRUE)
```

### Scatter Plot

Used for two quantitative variables.

```{r scatterplot-math-verbal-scores}
plot(schDat$Math_Score[schDat$Grade == 8], schDat$Verbal_Score[schDat$Grade == 8], xlab = "Math Score", ylab = "Verbal Score")
```

### Correlation Coefficient

Measures linear association.

```{r dat-des-linear-correlation}
cor(schDat$Math_Score, schDat$Verbal_Score)
```

### Block Boxplots

Grouped boxplots.

```{r dat-des-grouped-boxplots}
boxplot(Math_Score ~ Income_Level, data = schDat)
```

### Block Scatter Plots

Scatter plots grouped by category.
