# Probability

## From Data Description to Probability

### Observed Data vs Underlying Process

In the previous section, we focused on describing data using tables, graphs, and numerical summaries.

However, an important limitation arises:

- the data we observe represent only a **finite collection of observations**
- these observations come from a **process that could generate many more outcomes**

For example, in the simulated school data:

- we observe a set of students  
- but we could have observed a different set of students  
- the values would change  

This suggests that the data are **one realization of a broader process**.

---

### The Idea of a Theoretical Distribution

```{definition, theoretical-distribution, name="Theoretical Distribution"}
A theoretical distribution is a mathematical description of all possible outcomes of a process and their associated likelihoods.
```

A theoretical distribution:

* describes all possible values a variable can take
* assigns a likelihood to each value
* is typically **not fully observable**

We only observe:

* a sample from this distribution

Our goal in statistics is to:

* understand this underlying distribution
* use it to describe and predict outcomes

---

### Example: Repeated Sampling Interpretation

To better understand this idea, imagine repeating a study many times under the same conditions.

Each time:

* we collect a new sample
* the observed values change
* patterns vary

However:

* all samples are generated from the same underlying distribution

This distribution represents:

* the long-run behavior of the process

---

```{r repeated-sampling-placeholder}
# KEEP your existing simulation or sampling code here
# Example: repeatedly sample from your school dataset or a model
```

---

### Deterministic vs Random Phenomena

It is useful to distinguish between two types of processes:

* Deterministic processes: outcomes are fully predictable
* Random processes: outcomes are uncertain

```{definition, random-phenomenon, name="Random Phenomenon"}
A random phenomenon is a process whose outcome cannot be predicted with certainty.
```

Most real-world processes in data analysis are random:

* student performance
* attendance
* test scores

Probability provides a framework to model this uncertainty.

---

## Probability as a Model for Uncertain and Complex Phenomena

### Why We Need a Model

In real-world settings:

* data are variable
* outcomes are uncertain
* we cannot observe all possibilities

Without a model:

* we cannot generalize beyond the observed data

Probability provides:

* a structured way to describe uncertainty
* a way to reason about variability

---

### Probability as a Distribution

Probability assigns values to outcomes:

* values between 0 and 1
* representing how likely each outcome is

These values define a **distribution**.

This distribution allows us to:

* understand patterns
* quantify uncertainty
* compare outcomes

---

### Connection to Data

It is essential to distinguish between:

* data: what we observe
* probability: the model for all possible outcomes

This relationship can be summarized as:

* Data = one realization
* Probability = full distribution

This idea will be critical when we move to:

* sampling distributions
* statistical inference

---

## Random Experiments

```{definition, random-experiment, name="Random Experiment"}
A random experiment is a process that produces outcomes that cannot be predicted with certainty.
```

Examples:

* selecting a student at random
* recording a test score
* counting the number of absences

---

## Sample Space and Events

### Sample Space

```{definition, sample-space, name="Sample Space"}
The sample space is the set of all possible outcomes of a random experiment.
```

Examples:

* coin flip: ${H, T}$
* test score: all possible scores
* number of absences: ${0, 1, 2, \dots}$

---

### Events

```{definition, event, name="Event"}
An event is a subset of the sample space.
```

Examples:

* passing an exam
* scoring above 80
* having more than 5 absences

---

### Event Operations

We can combine events using:

* Union: $A \cup B$ (A or B)
* Intersection: $A \cap B$ (A and B)
* Complement: $A^c$ (not A)

---

```{r event-operations-placeholder}
# KEEP your existing code illustrating events and set operations
```

---

## Probability

### Definition of Probability

```{definition, probability, name="Probability"}
Probability assigns a number between 0 and 1 to each event, representing its likelihood.
```

---

### Basic Properties

For any event $A$:

* $0 \leq P(A) \leq 1$
* $P(S) = 1$

These properties ensure:

* probabilities are interpretable
* the total probability is 1

---

## Probability Rules

### Complement Rule

$$
P(A^c) = 1 - P(A)
$$

---

### Addition Rule (General)

$$
P(A \cup B) = P(A) + P(B) - P(A \cap B)
$$

---

### Addition Rule (Mutually Exclusive Events)

If $A \cap B = \emptyset$:

$$
P(A \cup B) = P(A) + P(B)
$$

---

```{r probability-rules-placeholder}
# KEEP your existing examples illustrating probability rules
```

---

## Conditional Probability

### Definition

```{definition, conditional-probability, name="Conditional Probability"}
The conditional probability of an event $A$ given $B$ is
$$
P(A \mid B) = \frac{P(A \cap B)}{P(B)}.
$$
```

---

### Multiplication Rule

$$
P(A \cap B) = P(A \mid B) P(B)
$$

---

```{r conditional-probability-placeholder}
# KEEP your existing conditional probability examples
```

---

## Independence

### Definition

```{definition, independence, name="Independence"}
Two events $A$ and $B$ are independent if
$$
P(A \cap B) = P(A)P(B).
$$
```

---

### Interpretation

Independence means:

* knowing one event occurs does not change the probability of the other

---

```{r independence-placeholder}
# KEEP your existing independence examples
```

---

## Random Variables

### Definition

```{definition, random-variable, name="Random Variable"}
A random variable is a function that assigns a numerical value to each outcome in the sample space.
```

---

### Discrete Random Variables

A discrete random variable takes:

* countable values

Examples:

* number of absences
* number of correct answers

---

### Continuous Random Variables

A continuous random variable takes:

* values in an interval

Examples:

* time spent studying
* exam scores (treated as continuous)

---

```{r random-variable-placeholder}
# KEEP your existing examples defining random variables
```

---

## Distributions of Random Variables

### Distribution Induced by a Random Variable

A random variable defines a distribution by assigning probabilities to its values.

---

### Probability Mass Function

For discrete variables:

$$
P(Y = y)
$$

---

### Probability Density Function

For continuous variables:

$$
f(y)
$$

---

### Cumulative Distribution Function

$$
F(y) = P(Y \leq y)
$$

---

```{r distribution-placeholder}
# KEEP your existing distribution examples
```

---

## Expected Value and Variance

### Expected Value

```{definition, expected-value, name="Expected Value"}
The expected value is the long-run average value of a random variable.
```

---

### Variance

```{definition, variance, name="Variance"}
Variance measures the variability of a random variable around its mean.
```

---

### Properties of Expectation

* Linearity:
  $$
  E(aY + b) = aE(Y) + b
  $$

---

### Properties of Variance

$$
\mathbb{V}(aY + b) = a^2 \mathbb{V}(Y)
$$

---

### Law of Total Expectation

```{definition, law-total-expectation, name="Law of Total Expectation"}
The expected value of a random variable can be computed by conditioning on another variable.
```

$$
E(Y) = E[E(Y \mid X)]
$$

---

### Linear Combinations

#### Expected Value of Linear Combinations

$$
E\left(\sum a_i Y_i\right) = \sum a_i E(Y_i)
$$

---

#### Variance of Linear Combinations (Independent Variables)

$$
\mathbb{V}\left(\sum a_i Y_i\right) = \sum a_i^2 \mathbb{V}(Y_i)
$$

---

#### Variance of Linear Combinations (Dependent Variables)

$$
\mathbb{V}(Y_1 + Y_2) = \mathbb{V}(Y_1) + \mathbb{V}(Y_2) + 2\text{Cov}(Y_1, Y_2)
$$

---

```{r expectation-variance-placeholder}
# KEEP your existing expectation/variance examples
```

---

## Important Probability Distributions

### Discrete Distributions

#### Binomial Distribution

```{r binomial-chunk}
# KEEP your original binomial code
```

---

#### Poisson Distribution

```{r poisson-chunk}
# KEEP your original poisson code
```

---

### Continuous Distributions

#### Uniform Distribution

```{r uniform-chunk}
# KEEP your original uniform code
```

---

#### Normal Distribution

```{r normal-chunk}
# KEEP your original normal code
```

---

#### Chi-Squared Distribution

```{r chisq-chunk}
# KEEP your original chi-squared code
```

---

#### t Distribution

```{r t-chunk}
# KEEP your original t code
```

---

#### F Distribution

```{r f-chunk}
# KEEP your original F code
```

---

## Summary

* Probability models a theoretical distribution
* Data are realizations from that distribution
* Random variables translate outcomes into numerical values
* Distributions describe variability
* Expectation and variance summarize key properties
* These ideas form the foundation for statistical inference