# Probability

## From Data Description to Probability

### Observed Data vs Underlying Process

In the previous section, we described data using numerical and graphical summaries.

However, the data we observe are only a **subset of all possible outcomes** of a process.

In most applications, there is an underlying mechanism generating the data, and what we observe is only one realization of that process.

---

### The Idea of a Theoretical Distribution

```{definition, theoretical-distribution, name="Theoretical Distribution"}
A theoretical distribution is a mathematical description of all possible outcomes of a process and their associated likelihoods.
```

* We do not observe the full distribution
* We only observe a sample from it
* Our goal is to model this underlying structure

---

### Example: Repeated Sampling Interpretation

If we repeatedly collected data under the same conditions:

* we would obtain different samples
* each sample would vary
* all samples come from the same underlying distribution

The theoretical distribution represents this full set of possible outcomes.

---

### Deterministic vs Random Phenomena

* Deterministic phenomena: outcomes are fully predictable
* Random phenomena: outcomes are uncertain

Probability is used to model **random phenomena**.

---

## Probability as a Model for Uncertain and Complex Phenomena

### Why We Need a Model

Real-world processes are:

* variable
* uncertain
* partially observed

Probability provides a framework to describe this uncertainty.

---

### Probability as a Distribution

Probability assigns a value to each possible outcome, describing how likely it is.

This allows us to:

* model variability
* describe uncertainty
* make predictions

---

### Connection to Data

* Data = one realization
* Probability = full distribution

This connection is fundamental for statistical inference.

---

## Random Experiments

```{definition, random-experiment, name="Random Experiment"}
A random experiment is a process that produces outcomes that cannot be predicted with certainty.
```

---

## Sample Space and Events

### Sample Space

```{definition, sample-space, name="Sample Space"}
The sample space is the set of all possible outcomes of a random experiment.
```

---

### Events

```{definition, event, name="Event"}
An event is a subset of the sample space.
```

---

### Event Operations

* Union: $A \cup B$
* Intersection: $A \cap B$
* Complement: $A^c$

---

## Probability

### Definition of Probability

```{definition, probability, name="Probability"}
Probability assigns a number between 0 and 1 to each event, representing its likelihood.
```

---

### Basic Properties

* $0 \leq P(A) \leq 1$
* $P(S) = 1$

---

## Probability Rules

### Complement Rule

$$
P(A^c) = 1 - P(A)
$$

---

### Addition Rule (General)

$$
P(A \cup B) = P(A) + P(B) - P(A \cap B)
$$

---

### Addition Rule (Mutually Exclusive Events)

If $A \cap B = \emptyset$:

$$
P(A \cup B) = P(A) + P(B)
$$

---

## Conditional Probability

### Definition

```{definition, conditional-probability, name="Conditional Probability"}
The conditional probability of an event $A$ given $B$ is
$$
P(A \mid B) = \frac{P(A \cap B)}{P(B)}.
$$
```

---

### Multiplication Rule

$$
P(A \cap B) = P(A \mid B)P(B)
$$

---

## Independence

### Definition

```{definition, independence, name="Independence"}
Two events $A$ and $B$ are independent if
$$
P(A \cap B) = P(A)P(B).
$$
```

---

### Interpretation

Independence means that knowing one event occurs does not change the probability of the other.

---

## Random Variables

### Definition

```{definition, random-variable, name="Random Variable"}
A random variable is a function that assigns a numerical value to each outcome in the sample space.
```

---

### Discrete Random Variables

A discrete random variable takes countable values.

---

### Continuous Random Variables

A continuous random variable takes values in an interval.

---

## Distributions of Random Variables

### Distribution Induced by a Random Variable

A random variable induces a distribution by assigning probabilities to its possible values.

---

### Probability Mass Function

For discrete variables:

$$
P(Y = y)
$$

---

### Probability Density Function

For continuous variables:

$$
f(y)
$$

---

### Cumulative Distribution Function

$$
F(y) = P(Y \leq y)
$$

---

## Expected Value and Variance

### Expected Value

```{definition, expected-value, name="Expected Value"}
The expected value is the average value of a random variable over many repetitions.
```

---

### Variance

```{definition, variance, name="Variance"}
Variance measures the variability of a random variable around its mean.
```

---

### Properties of Expectation

* Linearity:
  $$
  E(aY + b) = aE(Y) + b
  $$

---

### Properties of Variance

* Scaling:
  $$
  \mathbb{V}(aY + b) = a^2 \mathbb{V}(Y)
  $$

---

### Law of Total Expectation

```{definition, law-total-expectation, name="Law of Total Expectation"}
The expected value of a random variable can be computed by conditioning on another variable.
```

$$
E(Y) = E[E(Y \mid X)]
$$

---

### Linear Combinations

#### Expected Value of Linear Combinations

$$
E\left(\sum a_i Y_i\right) = \sum a_i E(Y_i)
$$

---

#### Variance of Linear Combinations (Independent Variables)

$$
\mathbb{V}\left(\sum a_i Y_i\right) = \sum a_i^2 \mathbb{V}(Y_i)
$$

---

#### Variance of Linear Combinations (Dependent Variables)

$$
\mathbb{V}(Y_1 + Y_2) = \mathbb{V}(Y_1) + \mathbb{V}(Y_2) + 2\text{Cov}(Y_1, Y_2)
$$

---

## Important Probability Distributions

### Discrete Distributions

#### Binomial Distribution

```{r binomial-chunk}
# KEEP your original binomial code chunk here
```

---

#### Poisson Distribution

```{r poisson-chunk}
# KEEP your original poisson code chunk here
```

---

### Continuous Distributions

#### Uniform Distribution

```{r uniform-chunk}
# KEEP your original uniform code chunk here
```

---

#### Normal Distribution

```{r normal-chunk}
# KEEP your original normal code chunk here
```

---

#### Chi-Squared Distribution

```{r chisq-chunk}
# KEEP your original chi-squared code chunk here
```

---

#### t Distribution

```{r t-chunk}
# KEEP your original t distribution code chunk here
```

---

#### F Distribution

```{r f-chunk}
# KEEP your original F distribution code chunk here
```

---

## Summary

* Probability models a theoretical distribution
* Data are realizations from that distribution
* Random variables translate outcomes into numerical form
* Distributions describe how values behave
* Expectation and variance summarize key features
* These ideas form the foundation for statistical inference
